Observations from Parallelising Three Maximum Common (Connected) Subgraph Algorithms

### 2017-08-13 paper/paper.tex
More things

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{cleveref}
\usepackage{tikz}

@@

?? Tree search. History.

@@

?? For the clique algorithm.

?? Discuss the TOPC paper.

?? Reproducible

@@

\begin{figure*}[tb]
    \includegraphics*{gen-graph-cumulative-plain.pdf}
    \hfill
    \includegraphics*{gen-graph-cumulative-33v.pdf}
    \hfill
    \includegraphics*{gen-graph-cumulative-33ved.pdf}

@@

In \cref{figure:cumulative} we plot empirical cumulative distribution functions showing the number
of instances solved over time, for both sequential (solid lines) and parallel (dotted lines)
versions of each algorithm.  Cumulative plots are commonly used in constraint programming and
boolean satisfiability for comparing aggregate behaviour of solvers (sometimes with the axes
reversed, in which case they are called \emph{cactus} plots).  To read these plots, make a choice of
timeout along the $x$-axis, which uses a log scale. The $y$ value at that point shows the number of
instances whose runtime (individually) is at most $x$, for a particular algorithm. In other words,
at any given $x$ value, the highest line shows which algorithm is able to solve the largest number
of instances using a per-instance timeout of that $x$ value, bearing in mind that the actual sets of
instances solved by each algorithm may be completely different.

@@

@@

\begin{figure*}[tb]
    \includegraphics*{gen-graph-scatter-plain-clique-vs-clique-par-t32.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-33v-clique-vs-clique-par-t32.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-33ved-clique-vs-clique-par-t32.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-plain-connected-clique-vs-clique-par-t32.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-33ve-connected-clique-vs-clique-par-t32.pdf}

@@

    \caption{On the top row, per-instance speedups, using the clique algorithm. The $x$-axis is
    sequential performance and the $y$-axis is 32 threaded performance, so points below the diagonal
    line represent a speedup. Darker points represent instances where the solution is relatively
    large compared to the order of the input graphs. Below, histograms plotting the distribution of
    speedups for instances whose sequential runtime was at least 500 milliseconds, and below the
    timeout.}
\end{figure*}

\begin{figure}[tb]
    \includegraphics{gen-graph-as-clique.pdf}

    \caption{Aggregate speedups from the clique algorithm using 32 threads, for each supported
    family of instances.}
\end{figure}

% \begin{figure*}[tb]
%     \includegraphics*{gen-graph-scatter-plain-clique-vs-clique-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-plain-kdown-vs-kdown-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-plain-mcsplit-vs-mcsplit-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-33v-clique-vs-clique-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-33v-mcsplit-vs-mcsplit-par-t32.pdf}
% 
%     \vspace*{1em}
% 
%     \includegraphics*{gen-graph-scatter-33ved-clique-vs-clique-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-33ved-mcsplit-vs-mcsplit-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-sip-kdown-vs-kdown-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-sip-mcsplit-vs-mcsplit-par-t32.pdf}
% 
%     \caption{Per-instance speedups. The $x$-axis is sequential performance and the $y$-axis is 32
%     threaded performance, so points below the diagonal line represent a speedup.}
% \end{figure*}
% 
\begin{figure}[tb]
    \includegraphics*{gen-graph-scatter-33ved-clique-vs-clique-par-t2.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-33ved-clique-par-t2-vs-clique-par-t4.pdf}

@@

\subsection{Old Text}

Cumulative performance plots comparing each parallel algorithm to its sequential version are given
in ??, using 32 threads on 16 core hyper-threaded Intel Xeon
E5-2640 v2 systems. The results are consistent: in each case the parallel algorithm is clearly and
substantially better than its sequential version, except when the sequential runtime is below 100
milliseconds.

?? gives instance-by-instance comparisons for a representative subset of
these algorithm and problem combinations. For the clique algorithm on unlabelled graphs (top left),
the results show fairly consistent speedups for harder instances: there is only a single instance
whose sequential runtime is over two seconds where we obtain a speedup of less than ten. We also see
a significant number of superlinear speedups. This is because there are relatively long proofs of
optimality for most of the harder instances, but strong solutions are usually found quickly (and
where they are not, the additional diversity helps). The results are similarly positive on
labelled graphs (centre left), with speedups of greater than ten being the norm for harder
instances, although there is more variety in the results.

For $k{\downarrow}$ on the unlabelled maximum common subgraph instances (top centre), the typical speedup
is between two and ten, and occasionally there are speedups of between ten and one hundred. A few
instances show small slowdowns: remember that when running 32 threads on a 16 core, hyper-threaded
system, each individual thread runs somewhat slower than it would if only a single thread were
running---in other words, these slight slowdowns are due to hardware effects, not changes to the
search tree.

On the larger subgraph isomorphism instances (bottom left), we see a concentration of speedups
of around ten. We also see strongly superlinear speedups in some cases, with eight
instances finishing in under one second which timed out sequentially at one thousand seconds.
Examining these instances closely shows that in each case, they are satisfiable with either $k = 0$
or very low values of $k$, and that the superlinear speedups are due to the parallel search
introducing early diversity, offsetting an incorrect initial value-ordering heuristic choice. We
also see more instances with slight absolute slowdowns: these are instances where the additional
threads contribute no helpful work to the solution, and where the slowdown due to not having 32
times as many resources when using 32 threads is particularly pronounced (memory contention is much
more of a problem for these larger graphs). Although not ideal, in aggregate these results are rare
and are more than offset by the superlinear speedups for other instances.

What about the splitting algorithm? On unlabelled instances (top right), our speedups are typically
between those obtained from the clique algorithm and those from $k{\downarrow}$, being mainly a
little below ten times. However, in some cases we see absolute slowdowns of up to four times: in
these cases, parallelism does little, and is not enough to offset the costs we paid when switching
from an in-place data structure to a data structure which requires copying. Interestingly, we also
see many more superlinear speedups for this algorithm, which suggests that its value ordering
heuristics are particularly poor early-on in search. (This last point is not particularly
surprising: compared to $k{\downarrow}$, the splitting algorithm does not try to reduce domains at
all at the top of search. The clique encoding, meanwhile, is able to capture even richer information
about variable-value relationships. The splitting algorithm is fast, but comparatively stupid.)

Similar trends occur with the subgraph isomorphism instances (bottom right), and the absolute
slowdowns go as high as a factor of ten---this is on instances with particularly large
target graphs, where copying is most expensive, and where there is a high branching factor at the
root of the search tree. Again, superlinear speedups are common.

On labelled instances (centre right), the parallel performance is less erratic. Although we no
longer see absolute slowdowns (excluding on very easy instances), we also see fewer superlinear
speedups, with most speedups being between one and ten. In many of these instances, parallelism
does very little: restricting splitting to a depth of five in these cases gives work balance
problems, but does help us avoid the substantial overheads that come with deeper splitting. This is
because the search trees for these instances tend to have a very low branching factor---in
constraint programming terms, the labels make each domain contain only a few values.

?? show that for all three
algorithms, parallel search is beneficial, and also (reasonably) risk-free: although in a few cases the
constant factor slowdowns due to hardware limitations and having to make changes to data structures
are unpleasant, we do not encounter any exponential slowdowns due to search tree changes. But what
about reproducibility and scalability? Further experiments in ??
establish that both of these properties also hold. In the left-hand column, we plot each algorithm
run against itself, using 32 threads in both cases. For the clique and $k{\downarrow}$ approaches,
every instance is on the central diagonal line, showing an extremely high level of reproducibility.
For the splitting algorithm, there is a little more variability, particularly for easier instances:
inspecting the results suggests that this is not down to different amounts of search work being
performed, but rather whether or not helper threads end up triggering additional copying.

In the right-hand column of ??, we plot the effects of moving from 8
threads (on the same system) to 32 threads. Points below the diagonal line indicate an improvement.
For the clique approach, increasing the number of threads is unequivocally beneficial. For the
$k{\downarrow}$ algorithm, the benefits are slim, and sometimes performance worsens slightly due to
memory contention (not work done) but we do not see any exponential slowdowns. Finally, for the
splitting algorithm, the results are usually strictly better, although in a few cases the overheads
increase further giving a slowdown; we also see an increase in the occurrence of strongly
superlinear speedups.

It is likely that the parallel results for the splitting algorithm can be improved further. In the
labelled case, we are often not obtaining a good work balance. However, deeper splitting is
expensive, particularly with larger graphs. One option to consider would be recomputing parts of the
search tree to avoid the costs of potentially allowing a copy in the future. This would be extremely
unpleasant from an implementation perspective, but may be necessary to obtain more even results.
However, even without this change, the parallel algorithm is clearly preferable to the sequential
one.

>>>>

\usepackage{showframe}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{nicefrac}

@@

The clique and $k{\downarrow}$ algorithms already make use of fine granularity bit-parallelism. To
introduce more coarse-grained thread parallelism, we will parallelise the search process: viewing
backtracking search as forming a tree, we can explore different portions of the tree on different
processors.

?? quick history

?? Anomalies

The maximum clique algorithm underlying . Differs only by a preprocessing step.

The clique-inspired connected algorithm described by \citet{} may be parallelised in exactly the
same way.

?? Note the very high number of recursive calls per second. Fitting everything in cache. Not
compatible with framework abstractions.

?? Splitting depth. Balance.

?? Discuss the TOPC paper.  ?? Reproducible, anomalies, etc

@@

@@

\begin{figure*}[p]
    \includegraphics*{gen-graph-cumulative-plain.pdf}
    \hfill
    \includegraphics*{gen-graph-cumulative-33v.pdf}
    \hfill
    \includegraphics*{gen-graph-cumulative-33ved.pdf}

@@

In \cref{figure:cumulative} we plot empirical cumulative distribution functions showing the number
of instances solved over time, for both sequential (solid lines) and parallel (dotted lines)
versions of each algorithm.  Cumulative plots are commonly used in constraint programming and
boolean satisfiability for comparing aggregate behaviour of solvers in the presence of timeouts.
(Sometimes the axes are reversed, in which case they are called \emph{cactus} plots.)  To read these
plots, make a choice of timeout along the $x$-axis, which uses a log scale. The $y$ value at that
point shows the number of instances whose runtime (individually) is at most $x$, for a particular
algorithm. In other words, at any given $x$ value, the highest line shows which algorithm is able to
solve the largest number of instances using a per-instance timeout of that $x$ value, bearing in
mind that the actual sets of instances solved by each algorithm may be completely different.

@@

However, although they are good at showing general trends, cumulative plots can hide interesting
details. We therefore now take a closer look at each of the three algorithms.

@@

\begin{figure*}[p]
    \includegraphics*{gen-graph-scatter-plain-clique-vs-clique-par-t32.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-33v-clique-vs-clique-par-t32.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-33ved-clique-vs-clique-par-t32.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-plain-connected-clique-vs-clique-par-t32.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-33ve-connected-clique-vs-clique-par-t32.pdf}

@@

    \caption{On the top row, per-instance speedups, using the clique algorithm. The $x$-axis is
    sequential performance and the $y$-axis is 32 threaded performance, so points below the diagonal
    line represent a speedup. Darker points represent instances where the solution is relatively
    large compared to the order of the input graphs. Below, histograms plotting the distribution of
    speedups for instances whose sequential runtime was at least 500 milliseconds, and below the
    timeout.}\label{figure:cliquespeedups}
\end{figure*}

In the top row of \cref{figure:cliquespeedups}, we see scatter plots comparing the sequential and
parallel runtimes of the clique algorithm on an instance by instance basis, using a log-log plot.
Instances which timed out using one algorithm but not the other are shown as points along the outer
axes. Points below the $x{-}y$ diagonal line represent speedups. The colour of the points indicates
the relative size of the solution---darker points represent instances where the solution uses most
of the vertices of the input graphs.

?? Quick summary

?? Try to turn this into a number, giving a speedup.

?? Best fit line. Pretend they are outliners. Can get away with it if they are rare enough, but they
are not for many algorithms.

?? Histograms. Ignore low runtimes. Also have to ignore instances solved in parallel but not
sequential. Underestimate superlinear speedups. We do not consider this to be a particularly
satisfactory way of measuring speedups.

?? We propose a different way of . Refer back to the cumulative plots. The usual way of reading
these plots is to pick whichever line is furthest up, and measuring the vertical difference between
lines says how many more the parallel algorithm can solve than the sequential algorithm can with a
particular choice of timeout. However, measuring the \emph{horizontal} distance between lines also
conveys information. Suppose the sequential algorithm can solve $y$ instances with a selected timeout
of $s$. By moving to the left on a cumulative plot, we can find the timeout $p$ required for the
parallel algorithm to solve the same number of instances, bearing in mind that \emph{the two sets of
instances could in principle have completely different members}. We define the \emph{aggregate
speedup} to be $\nicefrac{s}{p}$; this can be expressed as a function of time (i.e.\ $s$) or of the
number of instances solved ($y$).

\begin{figure}[p]
    \includegraphics{gen-graph-as-clique.pdf}

    \caption{Aggregate speedups from 32 threads on 16 hyper-threaded cores, shown as a function of
    sequential runtime, for each family supported by the clique algorithm.}
\end{figure}

% \begin{figure*}[tb]
%     \includegraphics*{gen-graph-scatter-plain-clique-vs-clique-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-plain-kdown-vs-kdown-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-plain-mcsplit-vs-mcsplit-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-33v-clique-vs-clique-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-33v-mcsplit-vs-mcsplit-par-t32.pdf}
% 
%     \vspace*{1em}
% 
%     \includegraphics*{gen-graph-scatter-33ved-clique-vs-clique-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-33ved-mcsplit-vs-mcsplit-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-sip-kdown-vs-kdown-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-sip-mcsplit-vs-mcsplit-par-t32.pdf}
% 
%     \caption{Per-instance speedups. The $x$-axis is sequential performance and the $y$-axis is 32
%     threaded performance, so points below the diagonal line represent a speedup.}
% \end{figure*}
% 
\begin{figure}[p]
    \includegraphics*{gen-graph-scatter-33ved-clique-vs-clique-par-t2.pdf}
    \hfill
    \includegraphics*{gen-graph-scatter-33ved-clique-par-t2-vs-clique-par-t4.pdf}

@@

% \subsection{Cilk}
%
% \begin{figure}[tb]
%     \includegraphics*{gen-graph-scatter-33ved-clique-par-t32r-vs-clique-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-33ved-clique-cilk-t32r-vs-clique-cilk-t32.pdf}
%
%     \vspace*{1em}
%
%     \includegraphics*{gen-graph-scatter-sip-kdown-par-t32r-vs-kdown-par-t32.pdf}
%     \hfill
%     \includegraphics*{gen-graph-scatter-sip-kdown-cilk-t32r-vs-kdown-cilk-t32.pdf}
%
%     \caption{Reproducibility using Cilk.}
% \end{figure}
